Pigeons.jl: Distributed sampling from intractable distributions

### 2024-05-29 paper/PT.tex
Fix bug
Tweaks
Minor tweaks

\subsection{Local exploration and communication}
A central data structure in Pigeons is the \texttt{Replica}, a Julia struct 
which we will often refer to in the remainder of this section. 
A single \texttt{Replica} stores a \texttt{state} variable and a \texttt{chain} 
integer, among other entries. 
At the beginning of PT, $N$ \texttt{Replicas} are created, 
one for each distribution on the annealing path, and 
the chain entries in the $N$ replicas are initialized at $1,2,\ldots,N$, respectively. 
For a given \texttt{Replica}, if the \texttt{chain} number is $i$ and the \texttt{state}
is $x$, then this means that the sample corresponding to the $i$-th distribution in 
the sequence $\pi_1, \ldots, \pi_N$ is currently at location $x$.

@@

\subsection{Distributed implementation}
A distributed implementation of PT is shown in \cref{alg:distributed_PT}
and we describe the details of the implementation below.
For pedagogy, we first present the algorithm in the special case where the number of machines available 
is equal to the number of \texttt{Replica}s and distributions, $N$. We outline 
how Pigeons removes this restriction in \cref{sec:relax}.
However, Pigeons.jl also allows for the more general case where the number of 
machines is not necessarily equal to $N$.

\subsubsection{The \texttt{PermutedDistributedArray}}
\label{sec:permuted_dist_array}
Recall that our theoretical $O(1)$ message size for communication between two machines 
is achieved by exchanging \texttt{chain} indices between \texttt{Replica}s 
instead of their \texttt{state}s. 
However to achieve this low cost a specialized distributed data structure is 
needed to maintain a mapping from chain indices to machine identifiers. 
To motivate it, we start with an illustrative example. 
Suppose we have $N=4$ distributions, chains, replicas, and machines, and that 
machine 1 is exploring chain 2 while machine 4 is exploring chain 3.
At one point, the \texttt{Replica} at \texttt{chain} 2 (machine 1) might need to exchange
\texttt{chain} indices with the \texttt{Replica} that has \texttt{chain} index 3 (machine 4). 
However, a priori it is not clear how machine 1 should know that it should communicate 
with machine 4 because it has no knowledge about the chain indices on the other machines.

@@

		  \If{$t$ is even} \Comment{choose between even or odd swap}
		    \State $\texttt{direction} \gets +1$
		  \Else
		    \State $\texttt{direction} \gets -1$
		  \EndIf
		  \State $\texttt{partner} \gets \texttt{chain} + \texttt{direction}$

@@

So far we have assumed, for pedagogy, that the number of 
machines $M$ is equal to the number of chains $N$.
This restriction is not present in Pigeons, so we now describe how 
the code base generalizes \cref{alg:distributed_PT}. 
First, the $N$ replicas are shared between the $M$ machines so that 
each machine contains either $\lceil N/M \rceil$ or $\lfloor N/M \rfloor$ 
replicas. Each machine's replicas are called the local replicas. 
The allocation of replicas to machine is fixed over the whole execution of the 
algorithm and maintained using the \texttt{LoadBalance} struct. 
Second, the code performing local exploration is enclosed in a 
loop over the local replicas, and if the options ``\texttt{multithreaded = true}'' 
is used in the \texttt{pigeons(...)} call, the exploration loop is performed in a 
\texttt{@threads for} loop. 
Finally, the functions used in the communication, 
\texttt{permuted\_get}, \texttt{permuted\_set} \texttt{transmit}, etc., 
are vectorized to accommodate that \texttt{chain} and 
\texttt{partner} are no longer integers but rather sets of integers. 
In particular, the permuted distributed array is designed so that each 
machine is responsible for tracking machine allocations for 
$\lceil N/M \rceil$ or $\lfloor N/M \rfloor$ chains instead of a single one.

@@

\begin{figure*}[t]
    \centering
    \begin{minipage}{0.4\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../img/speedup_distributed.pdf}
    \end{minipage}
    \begin{minipage}{0.4\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../img/speedup_parallel.pdf}
    \end{minipage}
    \caption{
        Weak scaling results for PT in both the distributed (MPI, {\bf left}) and parallel settings (Julia threads, {\bf right})  
        with approximate 95\% confidence intervals for the speedup relative to single-thread execution.   
        The blue dotted line represents a hypothetical speedup equal to the 
        number of processes/threads. 
        The red dashed line indicates a speedup of one for reference.
    }
    \label{fig:PT_scaling}
\end{figure*}

>>>>

\subsection{Local exploration and communication}
A central data structure in Pigeons is the \texttt{Replica}, a Julia struct 
that we will often refer to in the remainder of this section. 
A single \texttt{Replica} stores a \texttt{state} variable and a \texttt{chain} 
integer, among other entries. 
At the beginning of PT, $N$ \texttt{Replicas} are created, 
one for each distribution on the annealing path, and 
the chain entries in the $N$ replicas are initialized at $1,2,\ldots,N$, respectively. 
For a given \texttt{Replica}, if the \texttt{chain} number is $i$ and the \texttt{state}
is $x$, then this means that the sample corresponding to the $i$-th distribution in 
the sequence $\pi_1, \ldots, \pi_N$ is currently at location $x$.

@@

\subsection{Distributed implementation}
A distributed implementation of PT is shown in \cref{alg:distributed_PT}
and we describe the details of the implementation below.
For pedagogy, we first present the algorithm in the special case where the number of machines available 
is equal to the number of \texttt{Replica}s and distributions, $N$. 
However, Pigeons also allows for the more general case where the number of 
machines is not necessarily equal to $N$.
We outline how Pigeons achieves this in \cref{sec:relax}.

\subsubsection{Distributed mapping from chains to machines}
\label{sec:permuted_dist_array}
Recall that our theoretical $O(1)$ message size for communication between two machines 
is achieved by exchanging \texttt{chain} indices between \texttt{Replica}s 
instead of their \texttt{state}s. 
However, in order to achieve this low cost, a specialized distributed data structure is 
needed to maintain a mapping from chain indices to machine identifiers. 
To motivate it, we start with an illustrative example. 
Suppose we have $N=4$ distributions, chains, replicas, and machines, and that 
machine 1 is exploring chain 2 while machine 4 is exploring chain 3.
At one point, the \texttt{Replica} at \texttt{chain} 2 (machine 1) might need to exchange
\texttt{chain} indices with the \texttt{Replica} that has \texttt{chain} index 3 (machine 4). 
However, a priori it is not clear how machine 1 should know that it should communicate 
with machine 4 because it has no knowledge about the chain indices on the other machines.

@@

		  \If{$\texttt{is\_even}(t) = \texttt{is\_even}(\texttt{chain})$} 
		    \State $\texttt{direction} \gets +1$
		  \Else
		    \State $\texttt{direction} \gets -1$
		  \EndIf
		  \State $\texttt{partner} \gets \texttt{chain} + \texttt{direction}$

@@

So far we have assumed, for pedagogy, that the number of 
machines $M$ is equal to the number of chains $N$.
This restriction is not present in Pigeons, so we now describe how 
the code base generalizes \cref{alg:distributed_PT}. 
First, the $N$ replicas are shared between the $M$ machines so that 
each machine contains either $\lceil N/M \rceil$ or $\lfloor N/M \rfloor$ 
replicas. Each machine's replicas are called the local replicas. 
The allocation of replicas to machine is fixed over the whole execution of the 
algorithm and maintained using the \texttt{LoadBalance} struct. 
Second, the code performing local exploration is enclosed in a 
loop over the local replicas, and if the option ``\texttt{multithreaded = true}'' 
is used in the \texttt{pigeons(...)} call, the exploration loop is performed in a 
\texttt{@threads for} loop. 
Finally, the functions used in the communication, 
\texttt{permuted\_get}, \texttt{permuted\_set!}, \texttt{transmit}, etc., 
are vectorized to accommodate that \texttt{chain} and 
\texttt{partner} are no longer integers but rather sets of integers. 
In particular, the permuted distributed array is designed so that each 
machine is responsible for tracking machine allocations for 
$\lceil N/M \rceil$ or $\lfloor N/M \rfloor$ chains instead of a single one.

@@

\begin{figure*}[t]
    \centering
    \begin{minipage}{0.4\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../img/speedup_distributed.pdf}
    \end{minipage}
    \begin{minipage}{0.4\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../img/speedup_parallel.pdf}
    \end{minipage}
    \caption{
        Weak scaling results for PT in both the distributed (MPI, {\bf left}) 
        and parallel (Julia threads, {\bf right}) settings 
        with approximate 95\% confidence intervals for the speedup relative to single-thread execution.   
        The blue dotted line represents a hypothetical speedup equal to the 
        number of processes/threads. 
        The red dashed line indicates a speedup of one for reference.
    }
    \label{fig:PT_scaling}
\end{figure*}

### 2024-05-29 paper/SPI.tex
Minor tweaks

\medskip 
To avoid the issue of non-associativity of 
floating point arithmetic, we ensure that the order in which operations are performed 
is exactly the same, irrespective of the number of processes/machines and threads. 
This is achieved by making sure that every value to be added---if addition is our 
reduction operation---is a leaf node in the reduction tree, irrespective of the 
number of machines available to perform the reduction. 
For instance, if $N$ values are to be reduced, then the reduction tree would have 
$N$ leaf nodes. If $M$ machines are available, these machines are then assigned 
in such a way that the order of operations is as if there were $N$ machines available. 
To do so, it may be necessary for a machine to ``communicate with itself'', 
imitating the behaviour that would be present if there were $N$ machines available.
\cref{fig:reduction_tree_colour_8} and \cref{fig:reduction_tree_colour_2} 
illustrate the reduction procedure for eight and two machines, respectively.

>>>>

\medskip 
To avoid the issue of non-associativity of 
floating point arithmetic, we ensure that the order in which operations are performed 
is exactly the same, irrespective of the number of processes/machines and threads. 
This is achieved by making sure that every value to be added---if addition is our 
reduction operation---is a leaf node in a reduction tree that is invariant to the 
number of machines available to perform the computation. 
For instance, if $N$ values are to be reduced, then the reduction tree would have 
$N$ leaf nodes. If $M$ machines are available, these machines are then assigned 
in such a way that the order of operations is as if there were $N$ machines available. 
To do so, it may be necessary for a machine to ``communicate with itself'', 
imitating the behaviour that would be present if there were $N$ machines available.
\cref{fig:reduction_tree_colour_8} and \cref{fig:reduction_tree_colour_2} 
illustrate the reduction procedure for eight and two machines, respectively.

### 2024-05-29 paper/introduction.tex
Tweaks
Minor tweaks

\medskip
Pigeons.jl\footnote{The source code for Pigeons.jl can be found at 
\url{https://github.com/Julia-Tempering/Pigeons.jl}. 
Pigeons.jl v0.2.0, used in the example scripts below, is currently compatible with Julia 1.8+.}
enables users to sample efficiently from challenging probability distributions 
and solve integration problems by 
implementing state-of-the-art sampling algorithms \cite{syed2021nrpt,surjanovic2022vpt} 
that leverage distributed computation. Its simple user interface facilitates performing such computations 
single-threaded, multi-threaded, and/or distributed over MPI-communicating machines. 
We have stress-tested Pigeons successfully with up to 1000 machines running 
concurrently on a compute cluster (Sockeye) at the University of British Columbia.
Further, Pigeons comes with guarantees on strong parallelism invariance wherein 
the output for a given seed is \emph{identical} irrespective 
of the number of threads or processes. 
Such a level of reproducibility is rare in distributed software but of 
great use for the purposes of debugging in the context of sampling algorithms, 
which produce stochastic output.
Specifically, Pigeons.jl is designed to be suitable and yield reproducible output for:
\begin{enumerate}
    \item one machine running on one thread;
    \item one machine running on several threads;
    \item several machines running, each using one thread, and
    \item several machines running, each using several threads.
\end{enumerate}

@@

More in-depth documentation and additional examples can be found at: \\ \url{https://pigeons.run/}.

\subsection{Problem formulation}
We describe the class of problems that can be approached using Pigeons.jl.
Let $\pi(x)$ denote a probability density
\footnote{This density may also be a probability mass function 
(discrete variables). We also allow for a combination of discrete/continuous variables.} 
called the \emph{target}. 
In many problems, e.g. in Bayesian statistics, the density $\pi$ is typically 
known only up to a normalizing constant, 
\[
\label{eq:normalizing_constant}
  \pi(x) = \frac{\gamma(x)}{Z}, \qquad Z = \int \gamma(x) \, \dee x,
\]  
where $\gamma$ can be evaluated pointwise but $Z$ is typically unknown.
Pigeons.jl takes as input the function $\gamma$.

@@

    \item When the user needs not only $\int f(x) \pi(x) \, \dee x$ but also
    the normalization constant $Z$. 
    Many existing tools focus on the former and struggle or fail to do the latter.

>>>>

\medskip
Pigeons.jl\footnote{The source code for Pigeons.jl can be found at 
\url{https://github.com/Julia-Tempering/Pigeons.jl}. 
Pigeons.jl v0.2.0, used in the example scripts below, is compatible with Julia 1.8+.}
enables users to sample efficiently from challenging probability distributions 
and solve integration problems by 
implementing state-of-the-art sampling algorithms \cite{syed2021nrpt,surjanovic2022vpt} 
that leverage distributed computation. Its simple user interface facilitates performing such computations 
single-threaded, multi-threaded, and/or distributed over MPI-communicating machines. 
We have stress-tested Pigeons successfully with up to 1000 machines running 
concurrently on a compute cluster (Sockeye) at the University of British Columbia.
Further, Pigeons comes with guarantees on strong parallelism invariance wherein 
the output for a given seed is \emph{identical} irrespective 
of the number of threads or processes. 
Such a level of reproducibility is rare in distributed software but of 
great use for the purposes of debugging in the context of sampling algorithms, 
which produce stochastic output.
Specifically, Pigeons.jl is designed to be suitable and yield reproducible output for:
\begin{enumerate}
    \item one machine running on one thread;
    \item one machine running on several threads;
    \item several machines running, each using one thread, and
    \item several machines running, each using several threads.
\end{enumerate}

@@

More in-depth documentation and additional examples can be found at: {\renewcommand{\UrlFont}{\ttfamily\normalsize}\url{https://pigeons.run/}}.

\subsection{Problem formulation}
We describe the class of problems that can be approached using Pigeons.jl.
Let $\pi(x)$ denote a probability density\footnote{This density may 
also be a probability mass function 
(discrete variables). We also allow for a combination of discrete/continuous variables.} 
called the \emph{target}. 
In many problems, e.g. in Bayesian statistics, the density $\pi$ is typically 
known only up to a normalizing constant, 
\[
\label{eq:normalizing_constant}
  \pi(x) = \frac{\gamma(x)}{Z}, \qquad Z = \int \gamma(x) \, \dee x,
\]  
where $\gamma$ can be evaluated pointwise but $Z$ is typically unknown.
Pigeons.jl takes as input the function $\gamma$.

@@

    \item When the user needs not only $\int f(x) \pi(x) \, \dee x$ but also
    the normalizing constant $Z$. 
    Many existing tools focus on the former and struggle or fail to do the latter.

### 2024-05-29 paper/paper.tex
Minor tweaks

\begin{abstract}

We introduce a software package, Pigeons.jl, that provides a way to leverage distributed computation to 
obtain samples from complicated probability distributions, such as 
multimodal posteriors arising in Bayesian inference and high-dimensional 
distributions in statistical mechanics. Pigeons.jl provides a simple user interface to 
perform such computations single-threaded, multi-threaded,
and/or distributed over thousands of MPI-communicating machines.
In addition, Pigeons.jl guarantees a property that we call strong parallelism invariance: the 
output for a given seed is identical irrespective of the number of threads and processes, 
which is crucial for scientific reproducibility and software validation. 
We describe the key features of Pigeons.jl and the approach taken to implement 
a distributed and randomized algorithm that satisfies strong parallelism invariance.

\end{abstract}

>>>>

\begin{abstract}
We introduce a software package, Pigeons.jl, that provides a way to leverage distributed computation to 
obtain samples from complicated probability distributions, such as 
multimodal posteriors arising in Bayesian inference and high-dimensional 
distributions in statistical mechanics. Pigeons.jl provides a simple user interface to 
perform such computations single-threaded, multi-threaded,
and/or distributed over thousands of MPI-communicating machines.
In addition, Pigeons.jl guarantees a property that we call strong parallelism invariance: the 
output for a given seed is identical irrespective of the number of threads and processes, 
which is crucial for scientific reproducibility and software validation. 
We describe the key features of Pigeons.jl and the approach taken to implement 
a distributed and randomized algorithm that satisfies strong parallelism invariance.
\end{abstract}
